---
title: "Day 7 - part 1"
output:
  html_document:
    df_print: paged
---

```{r, message = FALSE}
library(tidyverse)
```

# Intcode computer

`exec_instr` executes an instruction from program `d` at postion `ptr` and takes input from `input` and returns the program with an updated `ptr` tacked onto the end of the program.

The computer `Intcode` takes the program, `program`, starting pointer position, `start_ptr`, and input, `input` and sequentially calls `exec_instr` until it runs into Opcode 99.

```{r Intcode}

Intcode <- function(program, input) {
ptr <- 0
input_ptr <- 1
output <- NULL
  
# Calling Intcode creates exec_instr and consequently exec_instr() will have Intcode parameter bindings in its surrounding scope, i.e. environment
# Alternatively create exec_instr in global environment and pass in as parameters values such as input
  
exec_instr <- function(ptr, d) {
  p <- ptr + 1
  Opcode <- d[p] %% 100
  para_modes <- 
    d[p] %/% (10 * 10^c(1:3)) %% (10^c(1:3)) # vector of parameter modes
  para_values <- 
    dplyr::if_else(para_modes == 0,
                   c(d[p + 1] + 1, d[p + 2] + 1, d[p + 3] + 1), # position mode 
                   c(p + 1, p + 2, p + 3) # immediate mode
    )
  
  if(Opcode == 1) {
    d[para_values[3]] <- d[para_values[1]] + d[para_values[2]]
    ptr <- ptr + 4
    return(c(d, ptr)) 
  } else if(Opcode == 2) {
    d[para_values[3]] <- d[para_values[1]] * d[para_values[2]]
    ptr <- ptr + 4
    return(c(d, ptr)) 
  } else if(Opcode == 3) {
    if(input_ptr <=2) {
    d[para_values[1]] <- input[input_ptr]
    print(paste("Taking input: ", input_ptr, " which as value ", input[input_ptr]))
    input_ptr <<- input_ptr + 1 # 
    ptr <- ptr + 2
    return(c(d, ptr))}
    else {
      
    }
  } else if(Opcode == 4) {
    output <<- d[para_values[1]]
    ptr <- ptr + 2
    # print(paste("Output: ", output, " at position ", p))
    return(c(d, ptr))
  } else if(Opcode == 5) {
      if(d[para_values[1]] != 0) ptr <- d[para_values[2]] 
      else ptr <- ptr + 3
    return(c(d, ptr)) 
  } else if(Opcode == 6) {
      if(d[para_values[1]] == 0) ptr <- d[para_values[2]]
      else ptr <- ptr + 3
    return(c(d, ptr)) 
  } else if(Opcode == 7) {
      if(d[para_values[1]] < d[para_values[2]]) d[para_values[3]] <- 1
    else d[para_values[3]] <- 0
    ptr <- ptr + 4
    return(c(d, ptr)) 
  } else if(Opcode == 8) {
      if(d[para_values[1]] == d[para_values[2]]) d[para_values[3]] <- 1
      else d[para_values[3]] <- 0
    ptr <- ptr + 4  
    return(c(d, ptr))       
  } else {
    print(paste("Invalid instruction at position ", p))
    # halt
    # Don't I want to return some output?  
    # And how about the state of the program?
  }
}

  pp <- length(program)
  program <- c(program, ptr)
  
  while(program[program[pp + 1] + 1] != 99) {
    program <- exec_instr(program[pp + 1], program[-(pp + 1)])
  }
  # print("Done")
  return(output)
}


```

# Day 5 TEST program

This computer will execute programs seen on day 2 and day 5.  

```{r}
TEST <- c(3,225,1,225,6,6,1100,1,238,225,104,0,1002,114,19,224,1001,224,-646,224,4,224,102,8,223,223,1001,224,7,224,1,223,224,223,1101,40,62,225,1101,60,38,225,1101,30,29,225,2,195,148,224,1001,224,-40,224,4,224,1002,223,8,223,101,2,224,224,1,224,223,223,1001,143,40,224,101,-125,224,224,4,224,1002,223,8,223,1001,224,3,224,1,224,223,223,101,29,139,224,1001,224,-99,224,4,224,1002,223,8,223,1001,224,2,224,1,224,223,223,1101,14,34,225,102,57,39,224,101,-3420,224,224,4,224,102,8,223,223,1001,224,7,224,1,223,224,223,1101,70,40,225,1102,85,69,225,1102,94,5,225,1,36,43,224,101,-92,224,224,4,224,1002,223,8,223,101,1,224,224,1,224,223,223,1102,94,24,224,1001,224,-2256,224,4,224,102,8,223,223,1001,224,1,224,1,223,224,223,1102,8,13,225,1101,36,65,224,1001,224,-101,224,4,224,102,8,223,223,101,3,224,224,1,223,224,223,4,223,99,0,0,0,677,0,0,0,0,0,0,0,0,0,0,0,1105,0,99999,1105,227,247,1105,1,99999,1005,227,99999,1005,0,256,1105,1,99999,1106,227,99999,1106,0,265,1105,1,99999,1006,0,99999,1006,227,274,1105,1,99999,1105,1,280,1105,1,99999,1,225,225,225,1101,294,0,0,105,1,0,1105,1,99999,1106,0,300,1105,1,99999,1,225,225,225,1101,314,0,0,106,0,0,1105,1,99999,8,677,226,224,1002,223,2,223,1006,224,329,1001,223,1,223,1108,226,226,224,1002,223,2,223,1005,224,344,101,1,223,223,1108,226,677,224,1002,223,2,223,1006,224,359,101,1,223,223,107,226,226,224,1002,223,2,223,1005,224,374,101,1,223,223,1107,226,226,224,1002,223,2,223,1005,224,389,101,1,223,223,107,677,677,224,102,2,223,223,1006,224,404,101,1,223,223,1008,226,226,224,1002,223,2,223,1006,224,419,101,1,223,223,108,677,226,224,1002,223,2,223,1006,224,434,101,1,223,223,1108,677,226,224,102,2,223,223,1005,224,449,101,1,223,223,1008,677,226,224,102,2,223,223,1006,224,464,1001,223,1,223,108,677,677,224,102,2,223,223,1005,224,479,101,1,223,223,7,677,677,224,102,2,223,223,1005,224,494,1001,223,1,223,8,226,677,224,102,2,223,223,1006,224,509,101,1,223,223,107,677,226,224,1002,223,2,223,1005,224,524,1001,223,1,223,7,677,226,224,1002,223,2,223,1005,224,539,1001,223,1,223,1007,226,677,224,1002,223,2,223,1005,224,554,1001,223,1,223,8,677,677,224,102,2,223,223,1006,224,569,101,1,223,223,7,226,677,224,102,2,223,223,1006,224,584,1001,223,1,223,1008,677,677,224,102,2,223,223,1005,224,599,101,1,223,223,1007,677,677,224,1002,223,2,223,1006,224,614,101,1,223,223,1107,677,226,224,1002,223,2,223,1006,224,629,101,1,223,223,1107,226,677,224,1002,223,2,223,1006,224,644,101,1,223,223,1007,226,226,224,102,2,223,223,1005,224,659,1001,223,1,223,108,226,226,224,102,2,223,223,1006,224,674,101,1,223,223,4,223,99,226)

Intcode(TEST, input = c(1, 1))
Intcode(TEST, input = c(5, 1))
```


# Two inputs

Enhance `Intcode` to take two inputs. See above.  Passed in a input vector and used the super-assignment operator `<<-` to increment an input vector.

# Amplifier Controller Software

First test `thruster` which is built on `Intcode` with two length 2 vector inputs and returning the last `output`.

```{r}
thruster <- function(ACT, phase_sequence) {
  A <- Intcode(ACT, c(phase_sequence[1], 0)) 
  B <- Intcode(ACT, c(phase_sequence[2], A))
  C <- Intcode(ACT, c(phase_sequence[3], B)) 
  D <- Intcode(ACT, c(phase_sequence[4], C))   
  E <- Intcode(ACT, c(phase_sequence[5], D))
  return(E)
}

ACT <- c(3,15,3,16,1002,16,10,16,1,16,15,15,4,15,99,0,0)
phase_sequence <- c(4,3,2,1,0)
thruster(ACT, phase_sequence) # 43210
ACT <- c(3,23,3,24,1002,24,10,24,1002,23,-1,23,101,5,23,23,1,24,23,23,4,23,99,0,0)
phase_sequence <- c(0,1,2,3,4)
thruster(ACT, phase_sequence) # 54321
ACT <- c(3,31,3,32,1002,32,10,32,1001,31,-2,31,1007,31,0,33,1002,33,7,33,1,33,31,31,1,32,31,31,4,31,99,0,0,0)
phase_sequence <- c(1,0,4,3,2)
thruster(ACT, phase_sequence) # 65210 
```

# Permutations of the phase sequence

There are 5 factorial or 120 permutations of `0:4`.

```{r}
perm <- function(v) {
  n <- length(v)
  if (n == 1) v
  else {
    X <- NULL
    for (i in 1:n) X <- rbind(X, cbind(v[i], perm(v[-i])))
    X
  }
}

phase_sequence <- perm(0:4)
phase_sequence[1,]
phase_sequence[120,]
```

# Puzzle input and answer

```{r}
ACT <- c(3,8,1001,8,10,8,105,1,0,0,21,34,59,68,89,102,183,264,345,426,99999,3,9,102,5,9,9,1001,9,5,9,4,9,99,3,9,101,3,9,9,1002,9,5,9,101,5,9,9,1002,9,3,9,1001,9,5,9,4,9,99,3,9,101,5,9,9,4,9,99,3,9,102,4,9,9,101,3,9,9,102,5,9,9,101,4,9,9,4,9,99,3,9,1002,9,5,9,1001,9,2,9,4,9,99,3,9,1002,9,2,9,4,9,3,9,101,2,9,9,4,9,3,9,1001,9,2,9,4,9,3,9,101,1,9,9,4,9,3,9,102,2,9,9,4,9,3,9,1001,9,2,9,4,9,3,9,1001,9,2,9,4,9,3,9,1001,9,2,9,4,9,3,9,1001,9,2,9,4,9,3,9,102,2,9,9,4,9,99,3,9,1001,9,1,9,4,9,3,9,102,2,9,9,4,9,3,9,102,2,9,9,4,9,3,9,101,1,9,9,4,9,3,9,101,1,9,9,4,9,3,9,102,2,9,9,4,9,3,9,102,2,9,9,4,9,3,9,101,1,9,9,4,9,3,9,1001,9,2,9,4,9,3,9,1001,9,2,9,4,9,99,3,9,1002,9,2,9,4,9,3,9,102,2,9,9,4,9,3,9,102,2,9,9,4,9,3,9,1001,9,2,9,4,9,3,9,1001,9,2,9,4,9,3,9,102,2,9,9,4,9,3,9,101,1,9,9,4,9,3,9,1001,9,1,9,4,9,3,9,1002,9,2,9,4,9,3,9,102,2,9,9,4,9,99,3,9,101,1,9,9,4,9,3,9,1002,9,2,9,4,9,3,9,102,2,9,9,4,9,3,9,1002,9,2,9,4,9,3,9,102,2,9,9,4,9,3,9,1002,9,2,9,4,9,3,9,102,2,9,9,4,9,3,9,1002,9,2,9,4,9,3,9,101,1,9,9,4,9,3,9,101,2,9,9,4,9,99,3,9,1001,9,1,9,4,9,3,9,1001,9,2,9,4,9,3,9,101,1,9,9,4,9,3,9,102,2,9,9,4,9,3,9,1001,9,2,9,4,9,3,9,1002,9,2,9,4,9,3,9,101,1,9,9,4,9,3,9,1001,9,1,9,4,9,3,9,1001,9,2,9,4,9,3,9,1002,9,2,9,4,9,99)
phase_sequence <- perm(0:4)

thruster_output <- as.vector(120)
i <- 1 
while(i <= 120) {
  thruster_output[i] <- thruster(ACT, phase_sequence[i, ])
  cat(i)
  i <- i + 1
  cat(",")
}
max(thruster_output) # 70597

```

# PART 2 FEEDBACK LOOP

New phase parameters 5:9.  Apparently, amplifers keep asking for input. Somehow I need to pause the computer when that input isn't there.

```{r}
ACT <- c(3,26,1001,26,-4,26,3,27,1002,27,2,27,1,27,26,27,4,27,1001,28,-1,28,1005,28,6,99,0,0,5)
Intcode(ACT, c(5, 0))

```

